---
title: "Course Project" 
subtitle: "R Programming"
author: "Javier Fong - 100437994" 
output: 
  bookdown::html_document2:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
---
  
```{r setup, include=FALSE}
options(scipen = 10000)
library(rstudioapi)
setwd(dirname(getActiveDocumentContext()$path)) 
```

# Description

As my final project for the R Programming class I decided to analyze the *Ames Housing Data set* collected [here](https://www.kaggle.com/c/house-prices-advanced-regression-techniques/data?select=train.csv). 

The data set includes 81 features describing a wide range of characteristics of 1460 home in Ames, Iowa sold between 2006 and 2010. 

For the purpose of this task, we will limit the variables to the following list: 
  
Discrete variables: 
  
* YearBuilt - Original Construction Year
* Bedroom - Number of bedrooms 
* Kitchen - Number of Kitchens 
* TotRmsAbvGrd - Total number of rooms in the house 
* MoSold - Month of sale
* YrSold - Year of sale 

Continuous variables:
  
* LotArea - Lot size in Square Foot
* 1stFlrSF - First Floor square feet
* 2ndFlrSF - Second Floor square feet
* SalePrice - Amount of sale

Categorical Variables:
  
* LotConfig - Lot configuration
* HouseStyle - Style of dwelling
* Exterior1st - Exterior covering on house
* Heating - Type of heating

# Project 

## Libraries

I'll use the following packages to perform the analysis: 

* **dplyr**: Package for information manipulation.
* **reshape2**: Package for data shape manipulation.
* **caret**: Package for machine learning algorithms. CARET stands for **C**lassification **A**nd **RE**gression **T**raining. 
* **ggplot2**: Graphs and plot functions.  
* **moments** : Moments, kurtosis and skewness functions. 
* **ggmosaic** : Mosaic plot functions. 
* **gridExtra** : Plot arrangement functions. 

```{r, message=FALSE}
library(dplyr)
library(caret)
library(reshape2)
library(ggplot2)
library(moments)
library(ggmosaic)
library(kableExtra)
library(gridExtra)
library(waterfalls)
library(forcats)
library(caretEnsemble)

```

## Data load 

```{r}

dataset = read.csv("ames.csv")

dataset = dataset %>% 
  select(
    LotArea
    , LotConfig
    , HouseStyle
    , YearBuilt
    , Exterior1st
    , Heating
    , FrstFlrSF = X1stFlrSF
    , ScndFlrSF = X2ndFlrSF
    , BedroomAbvGr
    , KitchenAbvGr
    , TotRmsAbvGrd
    , MoSold
    , YrSold
    , SalePrice
  )

dataset = dataset %>% unclass() %>% as.data.frame(stringsAsFactors = T)
dataset %>% str()
dataset %>% summary()
```

## Discrete Variables

```{r, message=F}
disc_vars = dataset %>% 
  dplyr::select(
    YearBuilt
    , BedroomAbvGr
    , KitchenAbvGr
    , TotRmsAbvGrd
    , MoSold
    , YrSold
  )

disc_vars %>% 
  melt() %>% 
  group_by(variable) %>% 
  summarise(
    min_v = min(value)
    , max_v = max(value)
    , diff_values = n_distinct(value)
    , mode_v = which.max(tabulate(value))
  ) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  kbl(
    format = "html"
    , caption = "Discrete Variable Summary"
    , col.names = c(
      "Variable"
      , "Min"
      , "Max"
      , "# of Values"
      , "Mode")
  ) %>% 
  kable_styling()
```

With the above summary we get an idea of the nature of the discrete variables in our data set. All of them are discrete ordered variables, meaning they're are ordered by nature. One comes after the other. And with the exception of YearBuilt, all of them have very few distinct values. 


```{r, message=F, fig.cap="Discrete Variables Histogram", fig.align='center'}
disc_vars %>% 
  melt() %>% 
  ggplot(
    aes(x = value, fill = variable) 
  ) + 
  geom_histogram() + 
  facet_wrap(~variable, scale = "free") +
  theme(legend.position = "none")

```

Now with the help of the histogram plots above, we get a clear idea of the group size of each value in the variables. 

* For the yearBuilt, we see there was a peak in the years between 1950 and 1975, and a more recent one just after the 2000. 
* In the BedroomsAbvGr variable, we conclude there are 3 main values that contain most of the observations (2,3 and 4 bedrooms).
* In the KitchenAbvGr variable, as expected, almost all properties have just 1 kitchen with a few exceptions that have 2. 
* Examining the total number of rooms in the properties, we see it has a larger variability of possible values, but most of them are concentrated between 4 and 8. 
* For the month sold variable, we conclude that the months between may and august are the most frequent months when properties are sold in Ames. 
* As for the years of the sale, we have a similar number of observations for each year. This might be by design of the data set. 

## Continuous Variables 

```{r, message=F, fig.cap = "Countinuous Variable Histograms", fig.align= 'center'}
cont_vars = dataset %>% 
  dplyr::select(
    LotArea
    , FrstFlrSF
    , ScndFlrSF
    , SalePrice
  )

cont_vars %>% 
  melt() %>% 
  ggplot(
    aes(x = value, fill = variable)
  ) +
  geom_histogram() +
  geom_vline(
    data = cont_vars %>% 
      melt() %>% 
      group_by(variable) %>% 
      summarise(mean_v =  mean(value))
    , aes(xintercept = mean_v)
    , linetype = "dashed"
  ) +
  facet_wrap(~variable, scale = "free") +
  theme(
    legend.position = "none"
  ) + 
  ylab("Frequency") + 
  xlab("Value")
```

Using the histogram plot we can get a pretty good idea of the frequency of the values for each of  the continuous variables. We notice in the case of the **Lot Area** that almost all the houses have a similar size. We can say so given that a higher percentage of the observation accumulate around the mean line. That's not the case with the **First Floor SqrFt** or the **Sale Price** variables, on which we notice higher dispertion in the values, even though most of the values tend to be smaller than the mean in both cases (we call this a left bias). Finally, it interesting to point out that with this plot we notice that most of the properties do not have a second floor. Notice how a high number of observation of the **Scnd Floor SqrFt** variable accumulate in the 0 value.  


```{r, message=F, fig.cap = "Countinuous Variable Boxplot", fig.align= 'center'}
cont_vars = dataset %>% 
  dplyr::select(
    LotArea
    , FrstFlrSF
    , ScndFlrSF
    , SalePrice
  )

cont_vars %>% 
  melt() %>% 
  ggplot(
    aes(x = value, fill = variable)
  ) +
  geom_boxplot() +
  facet_wrap(~variable, scale = "free") +
  theme(
    legend.position = "none"
  ) 
```

Now we use the help of boxplots of each variable to describe other characteristics. As expected from the histograms, we notice that the first and third quantile of the LotArea variable (Q1: 7,754, Q3: 11,602) area pretty close to each other, meaning that 50% of our observation are between these values. But we can see that we get a high number of outliers to the right side of the plot, meaning this houses are considerably larger (1.5x the IQR) than the average house. Another noticeable element of the plots is that the mean of the ScndFlrSF variable is pretty close to 0, supporting our observation that most properties do not have a second floor. The rest of the variables do not give us any remarkable new information. 

```{r, message=F}
cont_vars %>% 
  melt() %>% 
  group_by(variable) %>% 
  summarise(
    min_v = min(value)
    , max_v = max(value)
    , mean_v = mean(value)
    , sd_v = sd(value)
    , skweness_v = skewness(value)
    , kurtosis_v = kurtosis(value)
  ) %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  kbl(
    format = "html"
    , caption = "Continuous Variable Summary"
    , col.names = c(
      "Variable"
      , "Min"
      , "Max"
      , "Mean"
      , "Standard Dev."
      , "Skewness"
      , "Kurtosis")
  ) %>% 
  kable_styling()
```

Now we have a more precise values of our continuous variables using the above estimators. We'll interpret variable by variable: 

* In the case of the lot area, we get a mean value of $10,516.83 ft^2$ with a standard dev. of $9,981.26ft^2$. This sd seems to be influenced by the presence of the outliers observed in the boxplots. This variable also has the highest value of skewness (12.2), meaning most values are centered around the left tail of the distribution, as observed in the histogram. We also have an abnormally large value for the kurtosis in contrast with the rest of the variables. This high value means we're dealing with a *leptokurtic distribution*, meaning we have high concentration of data on the tails. 
* We can gather the same conclusions from the parameters of the FrstFlrSF and the SalePrice variables. The SD seems to indicate "relative" smaller dispersity in the observations with the mean. Although this variables have skewness and kurtosis values closer to 0 than the LotArea, they're still positive. Meaning most of our data are centered in the left tail of the distribution and a higher concentration of points in the tails than in the center. 
* The ScndFlrSF has the lower values for the skewness and kurtosis, meaning it might be the closest one resembling a normal distribution. Still, it is not a good match for it. 


## Categorical Variables

```{r, warning = F, message=F, fig.cap="Categorial Variables Distributions", fig.align='center'}
cat_vars = dataset %>% 
  dplyr::select_if(is.factor) %>% 
  dplyr::group_by_all() %>% 
  dplyr::summarise(cnt = n())



lotConfigPlot = cat_vars %>% ggplot(aes(x = fct_infreq(LotConfig), fill = LotConfig)) + geom_bar() + ylab("Frequency") + theme(legend.position = 'none') + xlab("LotConfig")
houseStylePlot = cat_vars %>% ggplot(aes(x = fct_infreq(HouseStyle), fill = HouseStyle )) + geom_bar()+ ylab("Frequency") + theme(legend.position = 'none')+ xlab("HouseStyle")
exteriorPlot = cat_vars %>% ggplot(aes(x = fct_infreq(Exterior1st), fill = Exterior1st)) + geom_bar()+ ylab("Frequency") + theme(legend.position = 'none',axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ xlab("Exterior1st")
heatingPlot = cat_vars %>% ggplot(aes(x = fct_infreq(Heating), fill = Heating)) + geom_bar()+ ylab("Frequency") + theme(legend.position = 'none')+ xlab("Heating")

grid.arrange(
  lotConfigPlot
  , houseStylePlot
  , exteriorPlot
  , heatingPlot
  , ncol = 2
)
```

From the above plots we can gather the following information: 

* LotConfig: 5 existing groups, but 2 (corner, inside) concentrate most of the information. 
* HouseStyle: 8 categories, with 3 considerably larger than the others. 
* Exterior1st: The most diverse variable. 15 group with 10 of them with considerable number of observations.
* Heating: 6 groups, but just 1 gather almost all the properties. 

```{r, fig.cap="Categorical Variables Accumulated Group Size %", fig.align='center'}
lotConfig_groupsize = cat_vars$LotConfig %>% 
  table() %>% 
  prop.table() %>% 
  sort(decreasing = T) %>% 
  data.frame() %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  waterfall(linetype = 1, fill_by_sign = F, fill_colours = 2:(nrow(.)+1)) + 
  geom_hline(yintercept = 0.8, linetype = "dashed") + 
  theme_minimal() + 
  xlab("LotConfig") + ylab("Accumulated Group Size %")

houseStyle_groupsize = cat_vars$HouseStyle %>% 
  table() %>% 
  prop.table() %>% 
  sort(decreasing = T) %>% 
  data.frame() %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  waterfall(linetype = 1, fill_by_sign = F, fill_colours = 2:(nrow(.)+1)) + 
  geom_hline(yintercept = 0.8, linetype = "dashed") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ 
  xlab("HouseStyle") + ylab("Accumulated Group Size %")

exterior_groupsize = cat_vars$Exterior1st %>% 
  table() %>% 
  prop.table() %>% 
  sort(decreasing = T) %>% 
  data.frame() %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  waterfall(linetype = 1, fill_by_sign = F, fill_colours = 2:(nrow(.)+1)) + 
  geom_hline(yintercept = 0.8, linetype = "dashed") + 
  theme_minimal()+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ 
  xlab("Exterior1St") + ylab("Accumulated Group Size %")

heating_groupsize = cat_vars$Heating %>% 
  table() %>% 
  prop.table() %>% 
  sort(decreasing = T) %>% 
  data.frame() %>% 
  mutate(across(where(is.numeric), round, 2)) %>% 
  waterfall(linetype = 1, fill_by_sign = F, fill_colours = 2:(nrow(.)+1)) + 
  geom_hline(yintercept = 0.8, linetype = "dashed") + 
  theme_minimal()+ 
  xlab("Heating") + ylab("Accumulated Group Size %")

grid.arrange(
  lotConfig_groupsize
  , houseStyle_groupsize
  , exterior_groupsize
  , heating_groupsize
  , ncol = 2
)
```

With the above waterfall plot we can know how many of the groups account for 80% of our observations. In the case of **LotConfig** we just need the 2 larger groups, *Inside* and *Corner* to group 76% of all the properties in our data set. In **HouseStyle** we need 4 out of the 8 groups to get 81% of the observations. The **Exterior1st**, as the most distributed variable, need 7/15 to sum up the 80% of the entries. At last we see that 84% of the observations in the **Heating** variable belong to the *GasA* category. 

## Analysis by Group 

```{r}
table(dataset$HouseStyle) %>% sort(decreasing = T )
```


Now, we'll analyze each group in the *HouseStyle* variable. From the previous section we know that the variable is divided in 8 different groups with the following distribution: 

* *1Story*: Properties with just 1 floor. 756 (28% of total) observations. 
* *2Story*: Properties with 2 floors. 445 obs (21%). 
* *1.5Fin*: Properties with the 2nd floor finished but unused. 154 obs (20%).
* *SLvl*: Split level. 65 obs (12%). 
* *SFoyer*: Splot foyer. 37 obs (8%). 
* *1.5Unf*: Properties with the 2nd floor under construction. 14 obs (~4%). 
* *2.5Unf*: Properties with the 3rd floor under construction. 11 obs (~4%). 
* *2.5Fin*: Properties with the 3rd floor finished but unused. 8 obs (~3%). 

```{r, fig.align = 'center', fig.cap="Continuous Variables Density by HouseStyle Group"}
hsdataset = dataset %>% 
  dplyr::select_if(names(.)=="HouseStyle" | sapply(., is.numeric))

hsdataset %>% 
  select(HouseStyle, LotArea, FrstFlrSF, ScndFlrSF, SalePrice) %>% 
  melt(id = c("HouseStyle")) %>% 
  ggplot(
    aes( 
      x = value 
      , col = HouseStyle
    )
  )  + 
  geom_density() + 
  facet_wrap(~variable, scale = "free")
```

First we take a look at the densities of the our continuous variables for each of the house style groups. For the lotArea and the ScndFlrSF variables we cannot get a clear view of any difference in the densities given the scale of the plots. Now for the FrstFlrSF variable, we do see some variation on each group. The *1.5Unf* sees to have the highest concentration of points near the $1000ft^2$. Then the rest of groups look like they have a more even distribution. It is interesting to notice that some of them, as *SLvl* and *2.5Fin* seem to be bimodal distributions, meaning there are 2 high concentration values of first floor area. For the Sale price, it is difficult to draw any conclusion from the densities of each group. It is still worth mentioning that there is a mode for *2.5Fin* group pretty far right in the plot. This is actually expected, given that this category is possibly the one with larger construction area, raising the price of this properties. 


```{r, fig.align = 'center', fig.cap="Continuous Variables Boxplot by HouseStyle Group"}
hsdataset %>% 
  select(HouseStyle, LotArea, FrstFlrSF, ScndFlrSF, SalePrice) %>% 
  melt(id = c("HouseStyle")) %>% 
  ggplot(
    aes( 
      x = value 
      , col = HouseStyle
    )
  )  + 
  geom_boxplot(varwidth = T) + 
  facet_wrap(~variable, scale = "free")
```

Now we use to the boxplots to get a more concise idea of the summary of the groups. Notice that the width of each box is directly proportional to the group size, meaning larger groups have wider boxes. Given the nature of our continuous variables, we should not expect to find any grand insight by house style group. Three out of the 4 variables measure floor area or the area on which the propertie is build. These variables are independent of the distribution of the house. 

That been said, the fourth variable, *SalePrice*, should have some relationship to the house distribution. In its boxplot we notice that the highest sold properties are those with 2 stories. Then we can compare properties with the same number of floors, but changing if they're still under construction. We see that those properties with finalized constructions (*1.5Fin* & *2.5Fin*) sell for highest prices than those with unfinished works (*1.5Unf* & *2.5Unf*) 


```{r, fig.align = 'center', fig.cap="Continuous Variables Scattered by HouseStyle Group and Year Built"}
hsdataset %>% 
  select(HouseStyle, YearBuilt, LotArea, FrstFlrSF, ScndFlrSF, SalePrice) %>% 
  melt(id = c("HouseStyle", "YearBuilt")) %>% 
  ggplot(
    aes( 
      y = value 
      , x = YearBuilt
      , col = HouseStyle
    )
  )  + 
  geom_point(size = 0.5) + 
  facet_wrap(~variable, scale = "free")


```

At last we tried to see if the there's been any interesting development though the years in this groups. We can see that the lot area has remained relatively the same for the last 100 years. But do notice that the construction of *1Story* houses has become more frequent in the last 50 years. Keeping in this same line, this *1Story* houses have seen an upward trend in its size sice the 1980's. We see the same pattern for the *2Story* houses, but with in more recent years. They become more often constructed in the early 2000's, and keep getting a larger second floor. At last we see an increment in sale price for all groups as the construction in more recent. 

## The CARET Package 

Now we'll use the caret package to make some predictions over our data set. 

### Model

#### Train and Test Set

First we'll divide our data set into a training and testing sets so we have a partition to train our model and a untouched one to test its performance. 

```{r}
set.seed(100)
train.ind = createDataPartition(
  dataset$HouseStyle
  , p = 0.8 
  , list = F
  , times = 1 
)

trainset = dataset[train.ind,]
testset = dataset[-train.ind,]

```

Our train and test set chosen at random have the following sizes: 

* Train Set: `r nrow(trainset)` rows.
* Test Set: `r nrow(testset)` rows.

#### Predictions 

##### Random Forest

```{r}
rf_fit = train(
  HouseStyle ~ . 
  , data = trainset 
  , method = "rf"
)

plot(
  varImp(rf_fit)
)

```
We build a random forest model to predict the *HouseStyle* base on all the other variables on our data set. From the above plot we notice that the most relevant variables the model found to predict this values are **ScndFlrSF**, **YearBuilt** and **FrstFlSF** above all others. 

Now  let's see how this model performs against unknown data. 

```{r}
confusionMatrix(
  reference = as.factor(testset$HouseStyle)
  , as.factor(predict(rf_fit, testset))
)
```

This model got an accuracy of 86% with a 95% confidence interval of 82% to 90%. That's an overall good model. But we can go even deeper and analize the performance of the model to predict each group. We see that the model is actually really good at finding 1Story of 2Story houses (high sensitivity for those categories). That's to be expected given that those are the 2 larger groups on our data set. On the contrary, the model is  pretty bad at finding properties that belong to all the other categories. For example, out of the 7 *SFoyer* properties in the test set, 6 got classified as *1Story* houses. The same with the *SLvl* group, where out of the 13 observations in the test set, 8 got classified as *1Story* and 3 as *2Story*. 

##### Naive Bayes

```{r, warning = F}
nb_fit = train(
  x = trainset %>% select(-HouseStyle)
  , y = trainset$HouseStyle
  , data = trainset 
  , method = "nb"
)

confusionMatrix(
  reference = as.factor(testset$HouseStyle)
  , as.factor(predict(nb_fit, testset))
)
```

```{r, warning=F}
svm_fit = train(
    HouseStyle ~ . 
  , data = trainset 
  , method = "svmLinear"
  , preProcess = c("center", "scale")
)

confusionMatrix(
  reference = as.factor(testset$HouseStyle)
  , as.factor(predict(svm_fit, testset))
)
```


#### Ensemble

```{r, warning = F}
trainControl = trainControl(
  method = "repeatedcv"
  , number = 5
  , repeats = 3
  , savePredictions = "final"
  , classProbs = T
)

algList = c("rf", "svmLinear")

models = caretList(
  HouseStyle ~ .
  , data = trainset 
  , trControl = trainControl()
  , methodList = algList
  , preProcess = c("center", "scale")
)

results = resamples(models)
summary(results)
```

